---
title: "Syn-bact_metagenomic_analysis"
output: html_notebook
---

Metagenomic workflow


steps

1. Pre-process (clean data)
2. Assembly via spades
3. orf predicton-prodigal
4. Functional annotation- Ghostkoala/kofamscan, Eggnog, Interproscan, Antismash (secondary metabolite biosynthesis gene), Cazy (carbohydrate)
5. Binning via metwrap and curation via anvio
6. Read abundance counting and TPM - featurecounts
7. Selective function exploration (N-P cycle, vitamin, host adhesion, QS/QQ, Bac. secretory system)


#1.0 Pre-processing (trim adapters, low quality reads and host contamination removal)

```{r}

conda activate sunbeam


#the samples should be in fq/fastq.gz form 
sunbeam init --data_fp /path/to/fastq/files /path/to/my_project

sunbeam init --data_fp ~/raw_files  ~/01.clean_reads/ --format {sample}.R{rp}.fq.gz

#This will create cofig.yml and sample files in project directory. Configure the configuration as per your needs. check for sample file (sample name should not contain hyphen, space. sample path should be absolute. sample name and path should match to actual files)

#Here we set configuration to clean adapters (no need to change), remove low quality reads and remove host contamination


sunbeam run -- --configfile ~/1.clean_reads/sunbeam_config.yml all_decontam -k -p --cores 20 --jobs 2 -w 90


```


#1.1 Fastqc

```{r}
#Check for sequence quality

mkdir -p ~/02.fastqc/

conda activate fastqc

#RAW reads

cd ~/sunbeam_output/qc/decontam/

fastqc Original.bottle.1.clean.R1.fastq.gz Original.bottle.1.clean.R2.fastq.gz Original.bottle.2.clean.R1.fastq.gz Original.bottle.2.clean.R2.fastq.gz suc.culture.100.days.1.clean.R1.fastq.gz suc.culture.100.days.1.clean.R2.fastq.gz suc.culture.100.days.2.clean.R1.fastq.gz suc.culture.100.days.2.clean.R2.fastq.gz suc.culture.250.days.1.clean.R1.fastq.gz suc.culture.250.days.1.clean.R2.fastq.gz suc.culture.250.days.2.clean.R1.fastq.gz suc.culture.250.days.2.clean.R2.fastq.gz suc.culture.400.days.1.clean.R1.fastq.gz suc.culture.400.days.1.clean.R2.fastq.gz suc.culture.400.days.2.clean.R1.fastq.gz suc.culture.400.days.2.clean.R2.fastq.gz suc.culture.450.days.1.clean.R1.fastq.gz suc.culture.450.days.1.clean.R2.fastq.gz suc.culture.450.days.2.clean.R1.fastq.gz suc.culture.450.days.2.clean.R2.fastq.gz -o path_to_directry/02.fastqc/ -t 80


```


#concatinate files 
```{r}
cd ~/sunbeam_output/qc/decontam/

cat *.clean.R1.fastq.gz > all_metag_conconated_R1.fastq.gz
cat *.clean.R2.fastq.gz > all_metag_conconated_R2.fastq.gz

Also concatinate technical replicates for assembly comparison (not shown here)

```


#3.0 Assembly 
```{r}

mkdir -p ~/03.assembly/


conda activate Spades

cd ~/sunbeam_output/qc/decontam/
  

spades.py --meta --pe1-1 all_metag_conconated_R1.fastq.gz --pe1-2 all_metag_conconated_R2.fastq.gz -t 50 -k 21,33,55,77,99,127 -o ~/03.assembly/
  

#check quality

conda activate quast

cd ~/03.assembly/
  
mkdir metaquast


python metaquast.py  --max-ref-number 0 --output-dir metaquast/scaffold.report scaffolds.fasta -t 30   
  
  

  
#3.1 Format scaffold file (convinient for downstream analysis with anvio)

conda activate anvio-7


mkdir -p ~/03.assembly/formatted_scaffold/



anvi-script-reformat-fasta  ~/03.assembly/scaffolds.fasta -o ~/03.assembly/formatted_scaffold/scaffolds-fixed.fa -l 0 --simplify-names

```


#4.0 Predict orf/cds
```{r}


mkdir -p ~/04.orf_predict


conda activate anvio-7


prodigal -i ~/scaffolds-fixed.fa -p meta -o ~/04.orf_predict/scaffolds-fixed.gff -a ~/04.orf_predict/scaffolds-fixed.faa -d ~/04.orf_predict/scaffolds-fixed.fa -f gff


```


#5.0 Functional annotation

```{r}

#5.1. GHOSTKOALA or Kofamscan (use any one)

#5.1.1 For GHOSTKOALA

#Due to size limitation of GHOSTKOAL (300mb), we split the protein sequence into two

conda activate pyfasta


pyfasta split -n 2 scaffolds-fixed.faa  
  
#(this will split the .fa sequence file into two)

#Upload it on Ghostkoala server (https://www.kegg.jp/ghostkoala/) for annotation


mkdir -p ~/05.Function/ghostkoala


#Once done, Transfer both the annotated ko files from ghostkoala to a directory and merge the two annotation file into one-

 

cat scaffolds-fixed.0_ko.txt scaffolds-fixed.1_ko.txt > scaffolds-fixed.ghostkoala_ko.txt



#5.2 EGGNOGG

mkdir -p ~/05.Function/eggnog


#Perform a 2-step (search + annotation) run, using Diamond in more-sensitive mode

conda activate eggnog-mapper

emapper.py --data_dir ~/database/eggnog/ -m diamond --sensmode more-sensitive --no_annot -i ~/04.orf_predict/scaffolds-fixed.faa -o ~/05.Function/eggnog/scaffolds-fixed_eggnog-no_annotation --cpu 30

#Then run,

emapper.py -m no_search --annotate_hits_table ~/05.Function/eggnog/scaffolds-fixed_eggnog-no_annotation.emapper.seed_orthologs -o scaffolds-fixed.eggnogg --output_dir ~/05.Function/eggnog/ --dbmem --cpu 30 


conda deactivate


#5.3 #2. INTERPROSCAN

mkdir -p  ~/05.Function/interproscan


####Interproscan requires the input files without any asterix (*), If your fasta file have asterix can remove by 
-

mkdir -p  ~/05.Function/interproscan/input/
  
  
cat ~/04.orf_predict/scaffolds-fixed.faa | perl -pe 's/\*//g' > ~/05.Function/interproscan/input/scaffolds-fixed.faa


#Now run interproscan,


~/interproscan-5.47-82.0/interproscan.sh -i ~/05.Function/interproscan/input/scaffolds-fixed.faa \
                    --output-dir ~/05.Function/interproscan/ \
--formats TSV, GFF3 \
  -cpu 30 \
  -iprlookup \
  -goterms \
  -pa \
  -dp



#5.4 ANTISMASH (For secondary metabolite annotation)

conda activate antismash



mkdir -p ~/05.Function/antismash/

antismash  ~/scaffolds-fixed.fa --cb-general --cb-knownclusters --cb-subclusters --asf --pfam2go --genefinding-tool none --genefinding-gff3 ~/04.orf_predict/scaffolds-fixed.gff --cpus 30 --output-dir ~/05.Function/antismash/ --verbose



#5.5 CAZZY

conda activate dbcan


mkdir -p ~/05.Function/acazy/
  
  

run_dbcan.py ~/04.orf_predict/scaffolds-fixed.faa protein  \
                  --dia_cpu 30 --hmm_cpu 30 --tf_cpu 30 \
                  --out_dir ~/05.Function/acazy/ \
                  --db_dir ~/database/dbcan2
                 
```


#6.0 Binning via metawrap

```{r}
#  6.1 Using metabat2,maxbin2 and concoct.


mkdir -p ~/06.binning/metawrap

conda activate metawrap


cd ~/sunbeam_output/qc/decontam/


metawrap binning -o ~/06.binning/metawrap -t 80 -a ~/scaffolds-fixed.fa --metabat2 --maxbin2 --concoct sample.clean_1.fastq.gz sample.clean_2.fastq.gz 

(replace sample in above command with the actual sample name. Mention all samples including the technical replicates)




# 6.1.1 Bin refinement
#using CheckM and min completion =50, max contamination=10
~/06.binning/metawrap/refined-bins -t 80 -A ~/06.binning/metawrap/metabat2_bins/ -B /~/06.binning/metawrap/maxbin2_bins/ -C ~/06.binning/metawrap/concoct_bins/ -c 50 -x 10



#6.2 ANVIO ( Here we use anvio to curate MAGS and summerize MAG info)


#anvi’o contigs database


mkdir -p ~/06.binning/anvio/01.contigDB  
  
conda activate anvio-7


anvi-gen-contigs-database -f ~/scaffolds-fixed.fa -o ~/06.binning/anvio/01.contigDB/contigs.db -n 'syn-bact-bins'



#6.3 Anvio's profile database


mkdir -p  ~/06.binning/anvio/02.profileDB  
 
cd ~/06.binning/metawrap/work_files/
 
 
dir > list.txt
#Edit the list to keep only the sample names
  
for SET in `cat list.txt`
do
anvi-profile -c /06.binning/anvio/01.contigDB/contigs.db \
-i $SET.clean.bam \
--num-threads 50 \
-o  ~/06.binning/anvio/02.profileDB/$SET
done



#This will create multiple profile db depending on number of samples. Now we merge these multiple db's into one

cd ~/06.binning/anvio/02.profileDB/


anvi-merge -c /06.binning/anvio/01.contigDB/contigs.db \
            100.days/PROFILE.db 250.days/PROFILE.db 400.days/PROFILE.db 450.days/PROFILE.db original.bottle/PROFILE.db \
            -o syn-bact-profiledb



#6.4 Import Ghostkoala annotation to anvio (optional)

#The  ghostkoala output gives genecalls and annotated kegg orthologs (KO). To add function to each KO we use kegg htext file


#download the kegg htext file (wget 'https://www.genome.jp/kegg-bin/download_htext?htext=ko00001&format=htext&filedir=' -O ko00001.keg) in the same directory containing the git cloned files (GhostKoalaParser).

mkdir -p ~/0.5functions/ghostkoala/GhostKoalaParser/
  
cd ~/0.5functions/ghostkoala/GhostKoalaParser/

wget 'https://www.genome.jp/kegg-bin/download_htext?htext=ko00001&format=htext&filedir=' -O ko00001.keg

#Then paste this command

kegfile="ko00001.keg"

while read -r prefix content
do
    case "$prefix" in A) col1="$content";; \
                      B) col2="$content" ;; \
                      C) col3="$content";; \
                      D) echo -e "$col1\t$col2\t$col3\t$content";;
    esac 
done < <(sed '/^[#!+]/d;s/<[^>]*>//g;s/^./& /' < "$kegfile") > KO_Orthology_ko00001.txt

#It will create a new file named KO_Orthology_ko00001.txt.





#Parse ko annotations to kegg functions



python ~/KEGG-to-anvio.py --KeggDB ~/GhostKoalaParser/KO_Orthology_ko00001.txt \
                          -i ~/ghostkoala/scaffolds-fixed.ghostkoala_ko.txt -o ~/ghostkoala/scaffolds-fixed.ghostkoala-KeggAnnotations-AnviImportable.txt


#Anvio expects gene_callers_id to be a numerical value. In our case wee need to remove prefix text from each gene_caller_id. To do so-

  
sed 's/c_//g' 0~/ghostkoala/scaffolds-fixed.ghostkoala-KeggAnnotations-AnviImportable.txt > ~/ghostkoala/scaffolds-fixed.ghostkoala-KeggAnnotations-AnviImportable1.txt


#Now import the new txt file with annotated functions into ANVIO. 



anvi-import-functions -c /06.binning/anvio/01.contigDB/contigs.db \
                        -i ~/ghostkoala/scaffolds-fixed.ghostkoala-KeggAnnotations-AnviImportable1.txt



#6.5 Import bin information into ANVIO using anvi-import-collection. metawrap will create this file for you.


anvi-import-collection --contigs-mode ~/06.binning/metawrap/refined-bins/metawrap_50_10_bins/metawrap4anvio.tsv -p ~/syn-bact-profiledb/PROFILE.db -c ~/01.contigDB/contigs.db --collection-name Metawrap_50_10



#6.5 Identifying single-copy core genes
 we will search our contigs for archaeal and bacterial single-copy core genes. This will be useful later on because
when we try to disentangle genomes from this metagenome, these single-copy core genes can be good markers for
how complete your genome is.

anvi-run-hmms -c ~/01.contigDB/contigs.db --num-threads 80



#optional functional annotation (recommended to compare with kegg annotations)


#NCBI's COGs (the database is not uptodate)

anvi-run-ncbi-cogs -c ~/01.contigDB/contigs.db --num-threads 80


#PFAM
anvi-run-pfams -c ~/01.contigDB/contigs.db --num-threads 80


#KEGG-KOFAM HMMS (needed if want to use anvio's metabolic pathway determination algorithm)

#GHOSTKOAL uses GHOSTX search (BLAST) against a nonredundant set of KEGG GENES while KOFAMSCAN usis HMM profiles.


anvi-run-kegg-kofams -c ~/01.contigDB/contigs.db \
-T 80 \
--collection-name Metawrap_50_10 \
-p ~/02.profileDB/syn-bact-profiledb/PROFILE.db 


#assign taxonomy (this is temporary taxonomy assignment. we will use gtdbtk for final taxonomy )


anvi-run-scg-taxonomy -c ~/01.contigDB/contigs.db -T 10 -P 5
                    

#Identify tRNAs (optional)
anvi-scan-trnas -c ~/01.contigDB/contigs.db --num-threads 50


#6.6 Identification and curation of MAGs

we use anvi-rename-bins to (1) filter  bins with >70% completion or >2 Mbp in size and at max 10% redundancy as metagenome-assembled genomes (MAGs), and (2) rename the MAGs and all the remaining bins based on their single-copy gene completion and redundancy values.


    anvi-rename-bins -c ~/01.contigDB/contigs.db \
                     -p ~/02.profileDB/syn-bact-profiledb/PROFILE.db \
                     --collection-to-read Metawrap_50_10 \
                     --collection-to-write Metawrap_70_10 \
                     --call-MAGs \
                     --min-completion-for-MAG 70 \
                     --max-redundancy-for-MAG 10 \
                     --prefix SN\
                     --report-file Metawrap_70-10_renamed_bins.txt


#6.7 SUMMERIZE MAGs RESULTS 


    anvi-summarize -c ~/01.contigDB/contigs.db \
                   -p ~/02.profileDB/syn-bact-profiledb/PROFILE.db \
                   -C Metawrap_70_10 \
                   -o  ~/06.binning/anvio/summary/SUMMARY-Metawrap_70_10-metagenomics \
                  --init-gene-coverages

    

Anvio will create a summary of the bins. Go through it. it will show important information about the bins like completion, redundancy, coverage, N50, Gene call, annotated functions, abundance etc.



#Determine metabolic pathway completion (via kegg-kofams annotations)



anvi-estimate-metabolism -c ~/01.contigDB/contigs.db -p ~/02.profileDB/syn-bact-profiledb/PROFILE.db -C Metawrap_70_10 



#6.8 Assign taxonomy to the bins

# copy all fasta (in anvio summary folder)  to a new directory

mkdir -p ~/06.binning/anvio/bin_taxo/fasta


cd ~/0.6anvio/

find summary/SUMMARY-Metawrap_70_10-metagenomics/bin_by_bin -name '*.fa' -exec cp -t ~/06.binning/anvio/bin_taxo/fasta {} +


# We will use gtdbtk package to assign taxonomy to our bins


conda activate gtdbtk-1.5.0

mkdir -p ~/06.binning/anvio/bin_taxo/gtdbtk


gtdbtk classify_wf --genome_dir ~/06.binning/anvio/bin_taxo/fasta --extension fa --out_dir bin_taxo/gtdbtk --cpus 30 --pplacer_cpus 20


#check taxonomic from above results


#6.9 Phylogenetic tree inference

#The _bac120.user_msa.fasta file from align folder in GTDBTk results is used for tree generation



#iqtree

mkdir -p ~/bin_taxo/iqtree

cp ~/bin_taxo/gtdbk/align/gtdbtk.bac120.user_msa.fasta ~/bin_taxo/iqtree


cd ~/bin_taxo/iqtree


conda activate iqtree2


iqtree2 -s gtdbtk.bac120.user_msa.fasta -st AA --alrt 1000 -B 1000 -T AUTO


#once finish you can use the generated tree in itol.


```



# 7.0 Mapping -Count read numbers of predicted orf
```{r}

mkdir -p  ~/06.binning/metawrap/7.featurecount 


cd ~/metawrap/work_files


conda activate subread


for SET in `cat list.txt`
do
featureCounts -T 30 -F gff -p -t CDS -g ID --verbose \
-a ~/scaffolds-fixed.gff \
-o ~/06.binning/metawrap/7.featurecount/$SET.featureCounts.txt \
$SET.clean.bam
done


# 7.1 Convert raw reads to tpm values

for SET in `cat list.txt`
do
Rscript ~/tpm_rpkm.R ~/06.binning/metawrap/7.featurecount/$SET.featureCounts.txt
done

```


# 7.2 Merging tpm file with functional annotation file
```{r}

#trim extra columns from the tpm file

mkdir -p ~/tpm

#Delete column 2-6
for SET in `cat list.txt`
do
cut -f1-1,7- ~/$SET.featureCounts_tpm.txt > ~/tpm/$SET.featureCounts_tpm.txt
done

#delete row-1
for SET in `cat list.txt`
do
sed -i '1d' ~/tpm/$SET.featureCounts_tpm.txt
done


#PARSING KO ANNOTATION TABLE (if the gene ids of the featurecount_tpm file and ghostkoala file are different, we need to make it matching)

#Rename genes in ko table TO MATCH WITH THE FEATURESCOUNTS

sed -r 's/^c_0*([0-9]*)/\1/' ~/scaffolds-fixed.ghostkoala_ko.txt > ~/scaffolds-fixed.ghostkoala_ko_1.txt


#now we merge the KO table with their TPM values

mkdir ~/06.Function/ghostkoala/ko_tpm/
  
#Use awk to map->>match gene-ids->>and make merged table

for SET in `cat list.txt`
do
awk 'NR==FNR { map[$1]=$2;next } { map1[$1]=$2 } END { PROCINFO["sorted_in"]="@ind_str_asc";for (i in map) { print i"\t"map[i]"\t"map1[i] } }' ~/ghostkoala/scaffolds-fixed.ghostkoala_ko_1.txt ~/tpm/$SET.featureCounts_tpm.txt > ~/ghostkoala/ko_tpm/$SET.ghostkoala_tpm.txt
done



#sort the data based on second column (ko)

mkdir -p ~/ko_tpm/sorted

for SET in `cat list.txt`
do
sort -k2 ~/ko_tpm/$SET.ghostkoala_tpm.txt > ~/ko_tpm/sorted/$SET.ghostkoala_tpm.txt
done

#Remove blank cells (gene_ids without kegg assignments):check and remove lines containing less than 3 strings 
for SET in `cat list.txt`
do
awk 'NF==3' ~/ko_tpm/sorted/$SET.ghostkoala_tpm.txt > ~/ko_tpm/sorted/$SET.ghostkoala_tpm1.txt
done

#Delete 1st column (gene_ids)

for SET in `cat list.txt`
do
 awk '{$1=""}1' ~/ko_tpm/sorted/$SET.ghostkoala_tpm1.txt > ~/ko_tpm/sorted/$SET.ghostkoala_tpm.txt
done 


#Now, Sum up the TPM OF multiple cells of same KO


mkdir -p ~/ko_tpm/sorted/summed


for SET in `cat list.txt`
do
awk '$1!=p{ if (NR>1) print p, s; p=$1; s=0} {s+=$2} END{print p, s}' ~/ko_tpm/sorted/$SET.ghostkoala_tpm.txt > ~/ko_tpm/sorted/summed/$SET.ghostkoala_tpm_summed.txt
done


#Do the same for cazy,antismash, eggnogg and interproscan annotations.

```


#8.0 Selective function exploration (important functions involved in  microbial interactions)
```{r}
#Here we want to find all the genes with their tpm values for specific functions. For that we need to create a list of ko's and their gene name for specific function/pathway
#First create a list of all ko and gene name for specific function. Then we map this list with the functional annotation (with tpm) file and extrect matching KOs


#For nitrogen cycle

mkdir -p ~/selective_funct/nitrogen_ko/

for SET in `cat list.txt`
do
awk 'FNR==NR { map[$1]=$2;next } { map1[$1]=$2 } END { PROCINFO["sorted_in"]="@ind_str_asc";for (i in map) { print i"\t"map[i]"\t"map1[i] } }' ~/nitrogen_ko.txt ~/ko_tpm/sorted/summed/$SET.ghostkoala_tpm_summed.txt > ~/selective_funct/nitrogen_ko/$SET-nitrogen.txt
done


#To join the resultant files

cd ~/selective_funct/nitrogen_ko/


join  original.bottle.1.nitrogen.txt original.bottle.2.nitrogen.txt| join - 100.days.1.nitrogen.txt | join - 100.days.2.nitrogen.txt | join - 250.days.1.nitrogen.txt | join - 250.days.2.nitrogen.txt | join - 400.days.1.nitrogen.txt | join - 400.days.2.nitrogen.txt | join - 450.days.1.nitrogen.txt | join - 450.days.2.nitrogen.txt > all.merged.nitrogen.txt



#Phosphorus cycle

mkdir -p ~/selective_funct/phosphorus_ko


for SET in `cat list.txt`
do
awk 'FNR==NR { map[$1]=$2;next } { map1[$1]=$2 } END { PROCINFO["sorted_in"]="@ind_str_asc";for (i in map) { print i"\t"map[i]"\t"map1[i] } }' ~/phosphorus_ko.txt ~/ko_tpm/sorted/summed/$SET.ghostkoala_tpm_summed.txt > ~/selective_funct/phosphorus_ko/$SET.phosphorus.txt
done


#Then join the files

#Repeat the above for vitamin, bacterial secretory system,quorum sensing and biofilm_formation gene sets.

```



